Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-28 Manuscript/manuscript.Rmd
manuscript edits

@@

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between regions with (minority class) and without (majority class) TAD boundaries, the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing].

In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. However, the class imbalance issue was not addressed. Even worse, when performing cross-validation [@mourad2016computational], used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. This is problematic in an imbalanced setting. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing]. For AUC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solution compare to simple random under-sampling, for instance random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role].

Furthermore, determing the optimal representation of the relationship between features in the model and genomic regions, known as feature engineering, plays a significant role in TAD prediction. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e. g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. Using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited out understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?] and technological [@Hsieh:2019aa] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relaticely coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).

Therefore, in this study, we aimed to develop an ensemble framework composed of different combinations of data driven characteristics used to address the issues presented above. We used our framework to systematically compare performances of random forest classifiers using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. This process was performed on chromosome-specific regions on the linear genome. Noninformative features were removed using a feature reduction algorithm and classifiers were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

@@

Biological replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained throughput the autosomal chromosomes. TAD boundaries called on Chromosome 9 were omitted from all downstream analyses, for both cell lines, due to the sparcity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries:

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a genomic bin overlapped with regions assigned to genomic annotations (features) was calculated. In the case of no overlaps, the value was set to 0.

@@

@@

Class imbalance was characterized by each dataset's imalance ratio, defined as the ratio of minority to majority classes given by the genomic bins. Minority classes were the set of genomic bins that overlapped with flanked TAD boundaries, while majority classes were the set of genomic bins that did not overlap with a flanked TAD boundary.

In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling), composed of the genomic bin class and data from each predictor type. The re-sampling techniques considered are described in detail below.

1. **No Sampling**: All of the data points from the majority and minority class of the training set were used.

2. **Random Under-Sampling (RUS)**: All of the minority classes from the training set were used. Sampling without replacement was used to obtain the same number of the majority classes.

3. **Random Over-Sampling (ROS)**: All of the majority classes from the training set were used. Sampling with replacement was used to obtain the same number of the minority classes.

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. The *DMwR* R package was used to perform SMOTE (version 0.4.1).

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 1). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. The cell line- and resolution-specific data was then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We analyzed publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were then identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line -specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 2).

@@

Expectedly, the median number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem was found to be much more consistant across resolutions in both cell lines. The sparsity of TAD boundaries throughout the genome was found to contribute to heavily imbalanced classes. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries ($Y=1$), to the much larger number of genomic bins that did not ($Y=0$). This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.
[??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."; !!! editted]

@@

We then investigated the relationship between the functional genomic annotations and the development of TAD boundaries by devising 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 3A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 3B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 3C).

@@

## DATA RE-SAMPLING COUPLED WITH THE INCLUSION OF A DISTANCE-TYPE FEATURE SPACE SUBSTANTIALLY IMPROVED MODEL PERFORMANCE

When models were built on randomly oversampled (ROS) data for GM12878, we found an increase in MCC compared to no resampling, at higher resolutions. The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Again, distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 3).

@@

Overall, SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. Here, distance type predictors were found to again yield the highest MCC values across resolutions. It should be noted that, SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 3). Moreover, for these sets of data characteristics, we again found that there was a degredation in model performance as data resolution decreased. Results indicated that performances exhibited much more variability when built on lower resolutions, specifically at the 100 kb level. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likwise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolution. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

@@

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary 10). 
# DISCUSSION

>>>>

**Methods.** To investigate the problem of spatial genomic relationships and class imbalance, we developed an ensemble framework for optimizing prediction of TAD boundaries. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) coupled with various metrics for classifier performance evaluation (Area Under Receiver Operating Characteristics (AUROC) curve, Precision-Recall curve (AUPRC), F1, Matthews Correlation Coefficient (MCC), among others) to select the best predictive model in class imbalance settings.

**Results.** Using different feature engineering and resampling strategies, coupled with elastic net-based feature selection, we systematically compared the performance of models predicting the location of TAD boundaries (Gm12878 and K562 Hi-C data at 10kb to 100kb resolutions). Precision-Recall curves were found to be best suited for evaluating model performance in class imbalance settings. Random undersampling and SMOTE-resampled data performed best for optimal parameter selection. Transcription Factor Binding Sites (TFBSs) were the best type of genomic annotations for predicting TAD boundaries when directly compared with histone modifications and chromatin states; we identified well-known CTCF, SMC3, RAD21, and several novel TFBSs as best predictors of TAD boundaries. Importantly, spatial relationships (distance between TAD boundaries and genomic annotations) were best predictors of TAD boundaries as compared with commonly used overlap counts or percent overlap measures.

@@

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between minority (regions with TAD boundaries) and majority classes (regions without TAD boundaries), the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing].

In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. However, the class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. This is problematic in an imbalanced setting. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing]. For AUC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling, for instance random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role].

Furthermore, determing the optimal representation of the relationship between features in the model and genomic regions, known as feature engineering, plays a significant role in TAD prediction. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e. g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. Here, using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. Moreover, while direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving forces of genome regulation [??? REFS]. These relationships can be quantified using distances (in basepairs) from regions on the linear genome to regions that define functional genomic annotations. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances have largely been ignored, yet have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

<!--
[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?] and technological [@Hsieh:2019aa] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).
-->

Therefore, in this study, we aimed to develop an ensemble framework composed of different combinations of data driven characteristics used to address the issues presented above. We used our framework to systematically compare performances of random forest classifiers using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution.

For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. This process was performed on chromosome-specific regions on the linear genome. Noninformative features were removed using a feature reduction algorithm. Lastly, random forest classifiers were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

@@

Biological replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained for the autosomal chromosomes. TAD boundaries called on Chromosome 9 were omitted from all downstream analyses, for both cell lines, due to the sparcity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries. They are described in detail below:

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a bin overlapped with regions assigned to genomic annotations (features) was calculated. In the case of no overlaps, the value was set to 0.

@@

### TRAINING AND TESTING

The cell line- and resolution-specific data was then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison.

@@

Class imbalance was characterized by each dataset's imalance ratio, defined as the ratio of minority to majority classes given by the genomic bins. Minority classes were the set of genomic bins that overlapped with flanked TAD boundaries, while majority classes were the set of genomic bins that did not overlap with a flanked TAD boundary. In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling). They are described in detail below.

1. **No Sampling**: All of the data points from the majority and minority classes of the training set were used.

2. **Random Under-Sampling (RUS)**: All of the data points from the minority class of the training set were used. Sampling without replacement was used to obtain the same number of data points from the majority class.

3. **Random Over-Sampling (ROS)**: All of the data points from the majority class of the training set were used. Sampling with replacement was used to obtain the same number of data points from the minority class.

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling [@chawla2002smote]. Under-sampling is performed without replacement from the majority class. On the other hand, over-sampling is performed by creating new synthetic observations using the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the feature vectors joining any/all of the k=5 (default) minority class nearest neighbors. The SMOTE algorithm can have many variations is the percentage of over- and under-sampling. In this study we restricted the algorithm to 100% over-sampling and 200% under-sampling. That is, 1 new synthetic observation was to be created for each existing minority class data point, while 2 data points were simultaneously randomly selected from the majority class. This combination results in perfectly balanced classes. The *DMwR* R package was used to perform SMOTE (version 0.4.1).

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 1). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We analyzed publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were then identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 2).

@@

Expectedly, the median number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem, caused by the sparsity of TAD boundaries throughout the genome, was found to be much more consistant across resolutions in both cell lines. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries, to the much larger number of genomic bins that did not. This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.

@@

To establish said feature space for all downstream random forest models, we constructed 3 different predictor-types. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 3A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 3B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 3C).

@@

## SMOTE RE-SAMPLING COUPLED WITH THE INCLUSION OF A DISTANCE-TYPE FEATURE SPACE SUBSTANTIALLY IMPROVED MODEL PERFORMANCE

When assessing performances for models built on balanced data, we found that SMOTE re-sampling provided the optimal MCC consistently across both predictor type and resolution. Specifically, the most improvement in model performance was seen when combining SMOTE re-sampling with distance-type predictors. This was also the optimal combination for the K562 cell line (Supplementary Figure 3). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. Moreover, there was a clear degredation in model performance as data resolution decreased. Likewise, performances exhibited much more variability across chromosomes when built on lower resolutions.

<!--
When models were built on randomly oversampled (ROS) data for GM12878, we found an increase in MCC compared to no resampling, at higher resolutions. The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Again, distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 3).

@@

Overall, SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. Here, distance type predictors were found to again yield the highest MCC values across resolutions. It should be noted that, SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 3). Moreover, for these sets of data characteristics, we again found that there was a degredation in model performance as data resolution decreased. Results indicated that performances exhibited much more variability when built on lower resolutions, specifically at the 100 kb level. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. 
-->

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likwise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

@@

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary 10).

# DISCUSSION
